---
title: "metabolic Genesets"
author: "Marie Kleinert"
date: "10 5 2022"
output:
  html_document: default
  word_document: default
---

Load all needed packages:

```{r, include = FALSE}
library(msigdbr)
library(pheatmap)
library(grid)
library(RColorBrewer)
library(bayesbio)
library(biomaRt)
genesets = readRDS("./data/hallmarks_genesets.rds")
ensembl = useMart("ensembl", dataset = "hsapiens_gene_ensembl")
```

Get human specific, canonical genesets from KEGG:

```{r}
canonical_pathways_KEGG = msigdbr(species = "human", category = "C2", subcategory = "CP:KEGG") 
```

Filtering for all genesets related to metabolism:

```{r}
metabolism_pathways_KEGG = canonical_pathways_KEGG[which(grepl("METABOLISM", canonical_pathways_KEGG$gs_name)==TRUE),]

rm(canonical_pathways_KEGG)
```

Creating a list with all characters for each pathway containing the human_gene_symbol of each related genes:

```{r}
metabolism_list_gs = list() #this creates an empty list to store the for loop in

for (gene.name in unique(metabolism_pathways_KEGG$gs_name))
  {pathway = c(metabolism_pathways_KEGG[metabolism_pathways_KEGG$gs_name == gene.name, 7])
  names(pathway) = substring(gene.name, 6) #to extract the "KEGG_" from each name, substring is used
  metabolism_list_gs[substring(gene.name, 6)] = pathway}

rm(gene.name, pathway) #this removes the selected vectors like pathways from the environment to keep it clearly arranged
```

Computing the Jaccard similarity for our list of gene sets "metabolism_list" and the given list of hallmark gene sets "hallmarks_genesets":

Creating a function that returns the Jaccard coefficient, considering that **Jaccard Similarity = (number of observations in both sets) / (number in either set)**

```{r}
jaccard = function(a, b) {                         #a und b sind Vektoren
    intersection = length(intersect(a, b))          #Schnittmenge
    union = length(a) + length(b) - intersection    #Vereinigungsmenge
    return (intersection/union)
}
```

Calculating the Jaccard-Coefficient for the metabolism genesets and the provided genesets of cancer hallmarks:

```{r}
Jaccard_coefficients = matrix(ncol = length(genesets[[1]]), nrow = length(metabolism_list_gs), byrow = TRUE) 

for (i in (1:length(metabolism_list_gs))){
  for (j in (1:length(genesets[[1]]))){
       j_value = jaccard(metabolism_list_gs[[i]], genesets[[1]][[j]])
    Jaccard_coefficients[i, j] = j_value 
  }} 

colnames(Jaccard_coefficients) = names(genesets[[1]])
rownames(Jaccard_coefficients) = names(metabolism_list_gs) 

rm(i, j, j_value)
```

Visualizing the similarity using a heatmap:

```{r}
pheatmap(Jaccard_coefficients, color = colorRampPalette(brewer.pal(n = 7, name ="PuBu"))(100), main = "Jaccard Coefficients of hallmark genesets and metabolism genesets", angle_col = "315", fontsize_row = 7.5, fontsize_col = 8)
```

Calculating the Jaccard-Coefficient for the metabolism genesets to each other:

```{r}
Jaccard_coefficients_met = matrix(ncol = length(metabolism_list_gs), nrow = length(metabolism_list_gs), byrow = TRUE) 

for (i in (1:length(metabolism_list_gs))){
  for (j in (1:length(metabolism_list_gs))){
       if (i == j){
         Jaccard_coefficients_met[i, j] = NA
       } else {
         j_value = jaccard(metabolism_list_gs[[i]], metabolism_list_gs[[j]])
    Jaccard_coefficients_met[i, j] = j_value 
  }}} 

colnames(Jaccard_coefficients_met) = names(metabolism_list_gs)
rownames(Jaccard_coefficients_met) = names(metabolism_list_gs) 

rm(i, j, j_value)
```

Visualizing the similarity using a heatmap:

```{r}
pheatmap(Jaccard_coefficients_met, color = colorRampPalette(brewer.pal(n = 7, name ="PuBu"))(100), main = "Jaccard Coefficients of metabolism genesets to each other", angle_col = "315", fontsize_row = 4.5, fontsize_col = 6)
```

Calculating the Jaccard-Coefficient for the provided genesets of cancer hallmarks to each other:

```{r}
Jaccard_coefficients_hall = matrix(ncol = length(genesets[[1]]), nrow = length(genesets[[1]]), byrow = TRUE) 

for (i in (1:length(genesets[[1]]))){
  for (j in (1:length(genesets[[1]]))){
       if (i == j){
         Jaccard_coefficients_hall[i, j] = NA
       } else {
         j_value = jaccard(genesets[[1]][[i]], genesets[[1]][[j]])
    Jaccard_coefficients_hall[i, j] = j_value 
  }}} 

colnames(Jaccard_coefficients_hall) = names(genesets[[1]])
rownames(Jaccard_coefficients_hall) = names(genesets[[1]]) 

rm(i, j, j_value)
```

Visualizing the similarity using a heatmap:

```{r}
pheatmap(Jaccard_coefficients_hall, color = colorRampPalette(brewer.pal(n = 7, name ="PuBu"))(100), main = "Jaccard Coefficients of hallmark genesets to each other", angle_col = "315", fontsize_row = 5.25, fontsize_col = 6)
```

Will using the ensembl ID change anything in the similarity compared to gene symbols?

Extracting the metabolism_list but with the human ensembl id of each gene instead of gene symbols:

```{r}
metabolism_list_id = list()

for (gene.name in unique(metabolism_pathways_KEGG$gs_name))
  {pathway = c(metabolism_pathways_KEGG[metabolism_pathways_KEGG$gs_name == gene.name, 9]) #just a different column from the canonical pathway dataframe that stores the ensembl id
  names(pathway) = substring(gene.name, 6) 
  metabolism_list_id[substring(gene.name, 6)] = pathway}

rm(gene.name, pathway) 
```

Converting the genesets list so it contains ensembl ids, as well:

```{r}
genesets_id = list() 

for (path in 1:length(genesets[[1]]))
{id = getBM(mart = ensembl, values = genesets[[1]][[path]], attributes = c("hgnc_symbol", "ensembl_gene_id"), filters = "hgnc_symbol")
  genesets_id[[path]] = id[[2]]} #path runs over the given geneset and with getBM we extract the ensembl id for each gene symbol of each pathway 

names(genesets_id) = names(genesets[[1]])

rm(path, id)
```

Calculating the jaccard coefficient between the metabolism genesets and the given hallmark genesets but both with the ensembl id:

```{r}
Jaccard_coefficients_id = matrix(ncol = length(genesets_id), nrow = length(metabolism_list_id), byrow = TRUE)
for (i in (1:length(metabolism_list_id))){
  for (j in (1:length(genesets_id))){
       j_value = jaccard(metabolism_list_id[[i]], genesets_id[[j]])
    Jaccard_coefficients_id[i, j] = j_value 
  }}

colnames(Jaccard_coefficients_id) = names(genesets_id)
rownames(Jaccard_coefficients_id) = names(metabolism_list_id)

rm(i, j, j_value, genesets_id, metabolism_list_id, metabolism_pathways_KEGG)
```

Plotting the similarity as a heatmap:

```{r}
pheatmap(Jaccard_coefficients_id, color = colorRampPalette(brewer.pal(n = 7, name ="PuBu"))(100), main = "Jaccard Coefficients of hallmark genesets and metabolism genesets with ensemble ID", fontsize_row = 5.25, fontsize_col = 6, fontsize = 8, angle_col = "315")

rm(Jaccard_coefficients_id, jaccard)
```
